### 2019-10-15 ch6_bred/content.tex
~

Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages\cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}. 
Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).
The case-wise GRNs -- or 'case-wise regulomes' -- are analysed similarly to single-cell transcriptomics data; for example by clustering, inferring trajectories, or finding differentially activated interactions.

@@

In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. 
\texttt{bred} uses case-wise feature importance\cite{xu_casespecificrandomforests_2016} score rather than the regular Random Forest feature importance scores. In addition to predicting regulatory strength of an interaction, \texttt{bred} also predicts the regulatory effect of an interaction -- that is, whether a regulator activates or represses the target.
We demonstrate \texttt{bred} by applying it on a single-cell dataset of 25'000 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 15'000 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.

@@

\begin{figure}[htb!]
	\centering
	\includegraphics[width=\linewidth]{fig/example.pdf} 
	\caption{
		\textbf{A schematic overview of the workings of the \texttt{bred} algorithm.}
	}
	\label{fig:example_cni}
\end{figure}

\begin{figure}[htb!]
	{\raggedright\textbf{A}}
	\begin{center}\includegraphics[width=.9\linewidth]{fig/tabula_muris_hema/grouped_interactions.pdf}\end{center}
	{\raggedright\textbf{B}}
	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
	\caption{
		\textbf{Hematopoietic cells from the Tabula Muris project.}
	}
	\label{fig:tabmur}
\end{figure}

\begin{figure}[htb!]
	{\raggedright\textbf{A}}
	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions.pdf}\end{center}
	{\raggedright\textbf{B}}
	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
	\caption{
		\textbf{The Cancer Genome Atlas.}
	}
	\label{fig:tcga}
\end{figure}

@@

@@

\subsection{Building the case-wise GRNs}
Having to infer a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely for every target $T$, predict which of the potential regulators regulate $T$ (Figure~\ref{fig:simplify}B). This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs. Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators. The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression. As in GENIE3, the target expression is first scaled to normalise feature importance scores across different targets.

@@

>>>>

Several studies have highlighted how some regulatory interactions can be very dynamic while others show evidence of being static during consecutive developmental stages\cite{moignard_characterizationtranscriptionalnetworks_2013, pina_singlecellnetworkanalysis_2015}. 
Since regulatory interactions are context-dependent\cite{papp_genomewideanalysiscontextdependence_2005}, attempting to create an accurate model of those processes by inferring a static regulatory network may have limited relevance.
Case-wise NI methods\footnote{Case-wise NI is sometimes also called sample-specific NI or case-specific NI.} avoid predicting a static GRN and instead infer one GRN per cell (or per sample, for bulk omics data).

@@

In this work, we introduce \texttt{bred}, the first 'true' case-wise NI method. For each interaction in the inferred case-wise GRN, \texttt{bred} predicts both the regulatory strength and its effect. 
The case-wise GRNs -- or 'case-wise regulomes' -- can be analysed analogously to transcriptomics data; for example by clustering samples, inferring trajectories, or finding differentially activated regions.
We demonstrate \texttt{bred} by applying it to a single-cell dataset of 22'122 hematopoietic cells from the Tabula Muris project\cite{schaum_singlecelltranscriptomics20_2018}, and to a collection of 14'963 bulk omics samples from The Cancer Genome Atlas project\cite{weinstein_cancergenomeatlas_2013}.

@@

\subsection{Hematopoietic cells from Tabula Muris}
From the Tabula Muris\cite{schaum_singlecelltranscriptomics20_2018} project, we selected all cells involved in some part of the hematopoietic lineage tree. We computed case-wise GRNs between 22'122 remaining single cells, for the 2000 most variable target genes and a subset of transcription factors as regulators.

To summarise the similarity in GRNs across cases, we first visualise a dimensionality reduction of the case-wise GRNs, where every dot represents a single GRN (Figure~\ref{fig:tabmur}A). The dimensionality reduction was computed by applying Fruchterman-Reingold\cite{fruchterman_graphdrawingforcedirected_1991} on the $k$-nearest-neighbour ($k$NN) graph of the case-wise GRN vectors. On the same $k$NN graph, the cells were clustered using the Louvain\cite{blondel_fastunfoldingcommunities_2008} clustering algorithm, and the clusters were labelled using prior information from Tabula Muris.
For each cluster, we retained the 50 interactions with the highest mean importance scores (Figure~\ref{fig:tabmur}B). In the visualisation, each node represents a gene, each edge an interaction, and the colour represents which cluster this interaction belongs to.

For the most part, clusters that are proximate in the dimensionality reduction (Figure~\ref{fig:tabmur}A) are also closely connected in the GRN network view (Figure~\ref{fig:tabmur}B). Notably, interactions from the different B cell clusters are almost not connected amongst each-other, while in the dimensionality reduction they re very proximate. Retaining more edges could result in the different B cells being connected in the GRN network view.

At the centre of the network lie many strongly connected interactions which are not specific for any particular cluster, but instead are common to almost all of the clusters. These are likely housekeeping genes or genes involved in key hematopoietic processes.

\begin{figure}[htb!]
	{\raggedright\textbf{A}}
	\begin{center}\includegraphics[width=\linewidth]{fig/tabula_muris_hema/plot_fr_cropped.png}\end{center}
	{\raggedright\textbf{B}}
	\begin{center}\includegraphics[width=.9\linewidth]{fig/tabula_muris_hema/grouped_interactions.pdf}\end{center}
	\caption{
		\textbf{22'122 case-wise GRNs of hematopoietic cells from the Tabula Muris project.}
		\textbf{A:} Each dot in this dimensionality reduction corresponds to the GRN of a single cell. The dimensionality of the cell-specific regulome matix was reduced using Fruchterman-Reingold and were clustered using Louvain clsutering, after which the clusters were relabelled using the meta information from Tabula Muris.
		\textbf{B:} Per cluster, the 50 interactions with the highest mean importance score are visualised.
	}
	\label{fig:tabmur}
\end{figure}

\subsection{The Cancer Genome Atlas}
We downloaded all available RNA-seq profiles from The Cancer Genome Atlas. In total, we computed case-wise GRNs for 14'963 tumour samples from 50 sub-projects, including 44 different cancer entities.

\begin{figure}[htb!]
	{\raggedright\textbf{A}}
	\begin{center}\includegraphics[width=.8\linewidth]{fig/tcga/plot_fr_cropped.png}\end{center}
	{\raggedright\textbf{B}}
	\begin{center}\includegraphics[width=.85\linewidth]{fig/tcga/grouped_interactions_2.pdf}\end{center}
	\caption{
		\textbf{14'963 case-wise GRNs of cancer bulk profiles from The Cancer Genome Atlas project.}
		\textbf{A:} The different profiles are visualised and clustered according to their regulomic similarities.
		\textbf{B:} Visualisation of the strongest interactions per cluster shows both pathways distinct to particular cancer entities as well as pathways common to multiple cancer entities.
	}
	\label{fig:tcga}
\end{figure}

@@

% going forward: functional validation; benchmark against in silico datasets; 
% can also use TI instead of clustering

@@

\subsection{Inferring case-wise GRNs}
The task of inferring a static GRN (Figure~\ref{fig:simplify}A) can be reduced to a simpler problem, namely: for every target $T$, predict which of the potential regulators $R_i$ regulate $T$ (Figure~\ref{fig:simplify}B). This simplification allowed GENIE3\cite{huynh-thu_inferringregulatorynetworks_2010} to use Random Forest's\cite{breiman_randomforests_2001} feature importance scores for inferring GRNs. Namely, a Random Forest is trained to predict the expression of a target gene of interest from the expression of potential regulators. The resulting Random Forest inherently allows to extract a feature importance score by observing the effect of each regulator in making a good prediction for the target expression. As in GENIE3, the target expression is first scaled to normalise feature importance scores across different targets.

@@

\subsection{Visualising clustered GRNs}
After Louvain clustering, the interactions of the 50 interactions with highest mean importance per cluster are retained. These interactions are visualised in Cytoscape\cite{shannon_cytoscapesoftwareenvironment_2003}, in which nodes depict genes, edges depict predicted regulatory interactions, coloured according to which cluster they are predicted for. The shape of the arrow denotes the predicted effect of the regulatory interaction.
